---
title: 'Human Activitiy Recognition : Course Project - Practical Machine Learning'
author: "courserans"
date: "March 19, 2015"
output:
  html_document:
    keep_md: yes
---

#### Background

Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively. These type of devices are part of the quantified self movement – a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it. 

The data set provided as part of the project exercise contains the accelerometer data for 6 participants who were asked to perform one set of 10 repetitions of the Unilateral Dumbbell Biceps Curl in five different manner. This is the "classe" variable in the training set.

- Class A : exactly according to the specification 
- Class B : throwing the elbows to the front
- Class C : lifting the dumbbell only halfway 
- Class D : lowering the dumbbell only halfway
- Class E : throwing the hips to the front

More information is available from the website here: http://groupware.les.inf.puc-rio.br/har (see the section on the Weight Lifting Exercise Dataset).

The goal of this project is to predict the manner in which a user did the exercise( Predict the "classe" variable). 

#### Load & Clean Data.

```{r, warning=FALSE}
library(caret)
library(randomForest)
library(plyr)
library(ggplot2)
library(corrplot)
library(gridExtra)
```

1. Load Data

```{r, cache=TRUE}
# Training : https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv 
# Test : https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv
test.raw <- read.csv("pml-testing.csv", na.strings=c("#DIV/0!","NA",""))
train.raw <-read.csv("pml-training.csv", na.strings=c("#DIV/0!","NA",""))
dim(train.raw)
```

2. Drop the rows with NA
3. Drop the first 7 columns which are user / observation specific markers and will not contribute to the prediction model.
4. Identify variables with zero variance so that predictor set can be reduced much further.
```{r}
predictors <- colnames(train.raw[colSums(is.na(train.raw)) == 0])[-(1:7)]
data <- train.raw[predictors]
nearZeroVar(data)
dim(data) ## 1 of the columns is classe and needs to be subtracted to arrive at the predictors.
```
As we note the predictor set has been reduced from **`r ncol(train.raw)`** to **`r ncol(data) - 1`**. Also the output of the nearZeroVar is zero which indicates that all the predictors have a non zero variance.

#### Exploratory Data Analysis

- Fig 1 from appendix : The user plot by classe indicates that there are 6 users and all of them have sufficient number of observation for each classe. 
- Fig 2 from appendix : The correlation plot indicates a number of variables with high correlations. We will not be removing these high correlation variables from the predictor, but instead allow the RF model to identify the most optimal predictor set.


#### Model Building
1. Create a test and training set
2. Fit a RF model with 5 fold cross validation
```{r, cache=TRUE}
set.seed(1021)
inTrain <- createDataPartition(y=data$classe, p=0.7, list=FALSE)
training <- data[inTrain,]
testing <- data[-inTrain,]

model <- train(classe ~ ., data=training, method="rf",
               trControl=trainControl(method="cv", 5))
model$finalModel
```

#### Model Results
```{r}
hat.testing <- predict(model,testing)
accuracy <- round(postResample(hat.testing, testing$classe)[1]*100,4)
error <- round((1 - as.numeric(confusionMatrix(testing$classe, hat.testing)$overall[1]))*100,4)
```
1. Fig 3 left panel of the plot lists the top 20 important features of the model. The roll_belt is the most important variable.
2. Mtry of 27 is the most optimal parameter for the model.
3. Accuracy of the model is **`r accuracy`%** and the out of sample error is **`r error`**%

#### Prediction

```{r}
hat.predict <- predict(model,test.raw)
hat.predict
```
Submit the files to coursera as per code block 1 from appendix.

#### Appendix

Fig 1. User by Classe

```{r, fig.align='center', fig.width=5, fig.height=4}
dat1 <- ddply(train.raw,.(classe,user_name),summarise,count=length(classe))
g1 <- ggplot(dat1,aes(x=factor(user_name),y=count, fill=classe)) + 
        geom_bar(stat="identity") + 
        scale_fill_brewer(palette=1) +
        xlab("User") + 
        theme_bw()
g1
```

Fig 2. Correlation plot

```{r, fig.align='center', fig.height=7, fig.width=7}
corrplot(cor(data[,-53])) ## 53 is the classe variable
```

Fig 3. Model Results

```{r, fig.align='center', fig.height=5, fig.width=8}
var.imp <- as.data.frame(varImp(model)$importance)
var.imp$feature <- row.names(var.imp)
g2 <- ggplot(head(var.imp,20),aes(x=reorder(feature,Overall),y=Overall)) + 
    geom_point() + coord_flip() + 
    ylab("Importance") + xlab("Predictors")+
    theme_bw()
grid.arrange(g2,plot(model),ncol=2)

```

Code 1 : Submission code.
```{r}

# pml_write_files = function(x){
#     n = length(x)
#     for(i in 1:n){
#         filename = paste0("problem_id_",i,".txt")
#         write.table(x[i],file=filename,quote=FALSE,row.names=FALSE,col.names=FALSE)
#     }
# }
# pml_write_files(hat.predict)
```